Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-23 MScDiplom_Lemenkova.tex
added new sections into ToC (List of Tools, Abstract, etc)

\textbf{Course Title:} \hfill{Geo-Information Science and Earth Observation for Environmental Modelling and Management}\\

@@

University of Southampton (UK)\\
Lund University (Sweden)\\
University of Warsaw (Poland)\\
University of Twente, Faculty ITC (The Netherlands)

@@

Thesis submitted to the University of Twente, faculty ITC, in partial fulfilment of the requirements
for the degree of Master of Science in Geo-information Science and Earth Observation for
Environmental Modelling and Management.\\

Thesis Assessment Board:\\

Name Examiner 1 (Chair)
Name Examiner 2
External Examiner
Member
Supervisors: Venus V. (Department of Natural Resource, ITC)
Toxopeus A.G. (Department of Natural Resource, ITC)\\

\begin{figure}[h]
	\begin{center}
	\includegraphics[height=2cm,width=8cm]{ITC-logo.jpg}
	\end{center}
\end{figure}

\begin{abstract}
\end{abstract}

\pagebreak
\textbf{Disclaimer}\\

This document describes work undertaken as part of a programme of study at the University of
Twente, Faculty ITC. All views and opinions expressed therein remain the sole responsibility of
the author, and do not necessarily represent those of the university.\\

\section*{}
\markright{Table of Content}
\addcontentsline{toc}{section}{Table of Content}
\tableofcontents

@@

\textbf{Software and utilities used in the current work}:\\
\begin{itemize}
	\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
	\item \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and classification
	\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water properties under changing environment
	\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and GPS for data capture during the fieldwork
	\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
	\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
	\item \href{http://www.python.org/}{Python language}, for writing auxiliary scripts (e.g. raw data interpolation)
	\item \href{http://www.gnuplot.info/}{Gnuplot} for plotting graphs
	\item \href{http://www.openoffice.org/}{Open Office} : document typesetting, data preprocessing and preliminary statistical evalution
	\item \href{http://calc2latex.sourceforge.net/}{Calc2LaTeX} : converting spreadsheets into \LaTeX environment
	\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)} : statistical data edition and analysis 
	\item \href{http://gretl.sourceforge.net/}{Gretl: Gnu Regression, Econometrics and Time-series Library} for statistic analysis
	\item \href{http://staruml.sourceforge.net/en/}{StarUML} , \href{http://live.gnome.org/Dia}{Dia} , \href{http://argouml.tigris.org/}{ArgoUML}: diagramming flowcharts for work algorithms
	\item \href{http://inkscape.org/}{Inkscape}: drawing
	\item \href{http://www.notepad-plus-plus.org/}{Notepad++}: auxiliary text editing and raw data formatting
\end{itemize}
\pagebreak

@@

\subsection{Research approach}
Seagrass consistent monitoring and mappint is necessary and important for the sustainable coastal
development and conservation measures. Earlier, many seagrass meadows have been destroyed by
human activities in the coastal zone, mainly due to the ignorance of their existence, because
information on the seagrass bed exact location was not available \cite{Choo06}\label{Choo06}. Well-time seagrass
observations and mapping enables precise control of its spatial distribution, detection of any changes
in the seagrass landscapes, highlightes potential environmental threats in the coastal zone (e.g.
declining of meadows) before they become unmanageable for the coastal management services.
Choosing the right and most effective approach method for the seagrass monitoring is essential.
Remote sensing methods alone, though having evident advantages, are insufficient, because satellite
images of underwater habitats are notoriously difficult to identify and interpret. The best research
method should be based on the integrated approach, well described in various scientific works \cite{Brown02,Montefalcone06,Kirkman96}\label{Brown02} \label{Montefalcone06} \label{Kirkman96}, which includes combination of various
techniques of the seagrass monitoring, i.e. remote sensing imagery classification of aerial and satellite
images, GIS-based spatial analysis and ground in-situ surveys.
The current study is based on the application of the remote sensing data, broadband satellite imagery,
aerial images and the results of the underwater videographic measurements towards seagrass
mapping. Image classification is based on the principle of the differentiation between the spectral
signatures of various seafloor cover types. The spectrum of light coming up from the ocean surface in
shallow waters keeps information on the optical properties of the seawater components and benthic
substrate which can be read from their spectral signatures \cite{Werdell03}\label{Werdell03}. \\

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.48]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}}
	\label{fig:10}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.6]{UML_GIS.png}
	\caption{General methodological research approach. \href{http://inkscape.org/}{Inkscape}}
	\label{fig:11}
\end{figure}

@@

\subsection{Seagrass global monitoring: history and perspectives}
Mapping and monitoring the seagrass is important for the environmental assessment of the marine
ecosystemsn in coastal areas. Regular tracking of current distribution of seagrass meadows, based on
correct information and cartographic visualization of seagrasses, is a preventive environmental
management, which helps to analyse potential environmental risks of coastal areas, decrease of the
number of species, loss of meadows and patches of the seagrasses.
The tradition of global seagrass mapping though has not a very long history comparing to the
terrestrial cartography, due to the technical difficulties of underwater observations, yet nowadays is
has become a successful, rapidly developping, increasingly popular and challenging research branch.
Seagrasses can be mapped using various approaches, with the most popular method of the remote
sensing application, enabling to detect and discriminate the substratum and various vegetation types
growing on the seafloor.
Several global seagrass surveys organise and provide regular monitoring of the segrass distribution,
health and species sustainability, amongst which the most world-known are, perhaps, the following
ones: \href{http://www.seagrassnet.org/}{Global Seagrass Monitoring project}, \href{http://www.seagrasswatch.org/}{Seagrasswatch}, 
located in Australia, \href{http://mediterranean.seagrassonline.org/}{the Mediterranean association Seagrass-2000},
\href{http://www.imedea.uib.es/index.php}{the Mediterranean Institute for Advanced Studies},
\href{http://www2.fiu.edu/~seagrass/}{Seagrass Ecosystems Research Laboratory in South Florida},
\href{http://www.flseagrass.org/}{Florida Seagrass organisation}, \href{http://www.seagrasses.org/}{the Seagrasses.org} and the 
\href{http://wsa.seagrassonline.org/}{World Seagrass Association}.
 These organizations aim at global seagrass monitoring, providing with
research results and reporting guidelines and manuals with standartized methods and
recommendations, specific for the seagrass research and monitoring. There are also university marine
centres and research institutes conducting seagrass monitoring and as a particular part of their
research. Their reports and guidelines were used for references in the current research.
Regular obserbvations and monitoring of the seagrasses are known since 1960s, mainly in tropical
regions (Australia). Since that time traditional methods of the seagrass monitoring and common
recommendations are being elaborated. The development of the underwater SCUBA diving
equipment and devices enabled to conduct underwater detailed measurements and observations
largely contributed to the improvement of the traditional in-situ observations of seagrass. From the
other side, development of the remote sensing methods and data acquisition from space contributed to
the new methods of seagrass mapping, using distance approach and generally based on images
classification.

@@

\begin{figure}
	\centering
	\subfloat [RAMSES-ACC-UV - Hyperspectral UVA/UVB Irradiance Sensor: 280-500 nm]{\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-2-1.jpg}}
	\subfloat [RAMSES-ARC - Hyperspectral UV-VIS Radiance Sensor: 320-950 nm.]{\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-2-2.jpg}}
	\caption{RAMSES Hyperspectral Sensors. Source: \href{http://www.trios.de}{Trios}}
	\label{fig:12}
\end{figure}

The optical properties of the sea water vary with different environmental conditions and reflect
current chemical content and physical specifics of the water, revealing the variability and distribution
of color of the sea waters, determined by the material in the water, e.g. chlorophyll, as well as its
physical properties, e.g. water absorption, attenuation, backscattering \cite{Maffione01a}\label{Maffione01a}.
Shallow waters generally contain more dissolved substances and suspended particles, which directly
influnces the transparency and colour of the waters of shelf zones \cite{Jerlov52}\label{Jerlov52}. Being highly
dynamic environments, coastal waters experience a variety of processes which alter their optical
properties incessantly. The effects of these processes influence application of the hyperspectral
remote sensing and reinforce other proceses \cite{Maffione01b}\label{Maffione01b}. Thus, waves and tides, increase
sedimentation processes which in turn may change microrelief properties and topology. 
The optical properties of the water are best reflected in the values of its radiance and irradiance,
which can be converted into spectral reflectance, or reflectivity. The spectral irradiance (E) is a
radiant flux of the electromagnetic solar radiation energy, received per surface unit area in a given
time (W·m−2·nm−1), while radiance (L) characterizes total emission or reflection that passes through
or is emitted from a particular area (W·sr−1·m−2 ·nm−1). Therefore, the spectral reflectance, or the
reflectivity of the object, can be estimated by the direct mathematical division of these first two
characteristics, and is expressed in persentage.
Therefore, the irradiance and radiance of the water should be measured in order to estimate spectral
reflectance of the various seafloor cover types. The optical measurements of the irradiance and
radiance of the sea water and bottom cover types of the seafloor can be received by the means of the
optical sensors spectroradiometers. There are several companies producing radiometers with various
characteristics and adjusted for defferent purposes, e.g. portable and miniature spectrometers from
\href{http://www.stellarnet.us}{StellarNet}, Hyperspectral Ocean Colour Radiometer (HyperOCR) sensor by
\href{http://www.satlantic.com}{Satlantic}, traceable spectroradiometers by \href{http://www.orboptronix.com}{Orbotronix} 
and lots of others. Among other radiometers there are ones designed by the \href{http://www.trios.de/}{Trios company} producing
optical sensors, GER-Series Field portable spectroradiometers from
\href{http://www.spectrapartners.nl/}{SpectraPartners}, etc. Trios-RAMSES hyperspectral radiometers
(Trios-RAMSES Hyperspectral UVA/UVB Irradiance Sensor and RAMSES-ARC Hyperspectral UV-
VIS Radiance Sensor) are small-sized, low power-consumpting, flexible for fieldwork yet with high
level of precision, specially calibrated for air and for water application as well as colour
measurements  (Fig.\ref{fig:10}). These products have been used for the radiance and irradiance
measurements in Agia Pelagia bay, Crete island, 2009.

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-12.jpg}
	\caption{Difference between broaddand multispectral and hyperspectral resolution of
		spectral signatures. Source: \cite{Clark99}\label{Clark99}}
	\label{fig:14}
\end{wrapfigure}

Seagrass meadows may reach spatial scales from several up to hundreds of metres, therefore they are
susceptible by the means of satellite imagery from remote sensors, both with moderate resolution
(e.g., \href{http://landsat.gsfc.nasa.gov/}{Landsat MSS, Landsat TM, Landsat ETM+}, \href{http://envisat.esa.int/instruments/meris/}{MERIS}, \href{http://asterweb.jpl.nasa.gov/}{ASTER}, \href{http://modis.gsfc.nasa.gov/}{MODIS}) and high resolution as
well (e.g., \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird}, \href{http://www.spotimage.fr/}{SPOT}, \href{http://www.itres.com/products/imagers/casi550}{CASI}). The possibility of their application towards seagrass mapping
vary and is limited by the technical characteristics (Table \ref{tab:1}) and resolution of these sensors  (Fig.\ref{fig:12}). In the next subparagraphs
we briefly discuss the limitations and research experience of the using of various imagery
for the seagrass mapping.

@@

\subsubsection{Review of multispectral imagery used for seagrass mapping}
Seagrass mapping using remotely sensed data from multispectral sensors is based on the classification
and discrimination of the seafloor types using their spectral characteristics in different wavebands.
Perhaps, the most known imagery, widely used in the remote sensing mapping, is received from the
group of Landsat sensors, known for its historical, pioneeer role in the satellite industry.
Being the longest running satellite system, lunched in 1972, Landsat is the only source of archival
data going back to 1984 at a sufficient spatial resolution \cite{Dekker05}\label{Dekker05}, which makes its data
desirable for historial mapping or environmental analysis of change detection of the seagrass
landscapes.
The Landsat TM and Landsat ETM+ data, with the most recent from sensor Landsat 7, an advanced
and multispectral scanning, launched in 1999, proved to be feasibile and useful for the mapping of
submerged vegetation, such as seagrasses or coral reefs. The successful applications of the imagery
Landsat TM towards the seagrass mapping, were reported in numerous research works \cite{Palandro03,Gullstroom06,Wabnitz08,Bierwirth93,Ferguson97,Rasib97}\label{Palandro03} \label{Gullstroom06} \label{Wabnitz08} \label{Bierwirth93} \label{Ferguson97} \label{Rasib97}. The Landsat data are particularly suitable for the
case of change detection of seagrass landscapes at a decadal scale, because being the main sensor
onboard the Landsat satellites, Landsat Thematic Mapper (TM) provides the longest time series
available for change detection analysis over submerged vegetation \cite{Palandro03}\label{Palandro03}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-13.jpg}
 	\caption{Band coverage of ASTER and Landsat channels on the e/m spectrum. Source:\cite{Kalinowski04}\label{Kalinowski04}}
	\label{fig:15}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-14.jpg}
	\caption{Multispectral vs. hyperspectral band coverage. Source: \cite{Sabins97}\label{Sabins97}}
	\label{fig:16}
\end{wrapfigure}

Advanced Spectrometer for Thermal Emission and Reflection Radiometer (\href{http://asterweb.jpl.nasa.gov/}{ASTER}), launched in
1999 onboard Terra sensor, provides high-resolution images of the Earth in 15 different bands of the
electromagnetic spectrum, ranging from visible to thermal infrared light  (Fig.\ref{fig:13}). It has diverse
subsystems for the visible near infrared (VNIR) with 15 m resolution, shortwave infrared (SWIR), and
thermal infrared (TIR) wavelength regions \cite{Kalinowski04}\label{Kalinowski04}. For each channel there is separate
onboard calibration (OBC) system, telescope with independent pointing and different detector technsology \cite{Arai05}\label{Arai05}. 
This makes ASTER imagery expecially suitable for the application towards detailed
mapping of surface temperature, emissivity and reflectance of objects and bathymetric elevations as
well. Successful application of the ASTER imagery towards mapping of submerged vegetation
reported, for example, by Hirose \cite{Hirose04} \label{Hirose04}. Other multispectra images have also been used for the
seagrass interpretation. For example, the success of using imagery from the multispectral airborne
scanner Daedalus AADS1268 is reported by Heege et al \cite{Heege03}\label{Heege03} where they aim at classification of
macrophytes in shallow waters of the Lake Constance.
In regard to the methods chosen for the image interpretation, the supervised classifications proves to
be the most worthy in a majority of case studies \cite{Palandro03,Peneva08}\label{Palandro03} \label{Peneva08}. While
often methods of the unsupervised classification are used as a tool for classifying submerged object
and features on the multispectral images and aerial imagery \cite{Fletcher09}\label{Fletcher09}, it still does not
provide common algorithms that can be applied to other images and regions. Thus, the strength of the
optical signal coming from the various types of the seafloor is strongly influenced by the effects of the
water column, its depth and chemical properties. Yet methods of the unsupervised classification
overleap mixed spectral effects of the water column which shift the real values of the spectra, with the
pure reflectance from the benthos; as a result, it may cause significant errors \cite{Dierssen03}\label{Dierssen03}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-15.jpg}
	\caption{Reflectance spectra of seagrass; Rrs(0-) - subsurface RS reflectance; Rrs b - the bottom
		reflectance. Source:\cite{Yang10}\label{Yang10}}
	\label{fig:17}
\end{wrapfigure}

Therefore, for the accurate assessments of various seafloor cover types, water
column depth and its optical properties, methods of supervised classification should be preferably
used as being more suitable in classification and interpretations of imagery.
While this and other studies \cite{Phinn08}\label{Phinn08}demonstrated the advantages and success of the
application of multispectral imagery for the spectral discrimination of the seafloor cover types and
mapping the submerged landscapes on the basis of pixels classification, the application of data from
hyperspectral sensors has better potential due to their higher resolution.

>>>>

\textbf{Course Title:} \\
{Geo-Information Science and Earth Observation for Environmental Modelling and Management}\\

@@

\begin{itemize}
	\item[*] University of Southampton (UK)\\
	\item[*] Lund University (Sweden)\\
	\item[*] University of Warsaw (Poland)\\
	\item[*] University of Twente, Faculty ITC (The Netherlands)
\end{itemize}
\pagebreak

@@

Thesis submitted to the University of Twente, Faculty ITC, in partial fulfilment of the requirements
for the degree of Master of Science in Geo-information Science and Earth Observation for
Environmental Modelling and Management.\\

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Name Examiner 1 (Chair)
	\item Name Examiner 2
	\item External Examiner
\end{itemize}

@@

\textbf{Member Supervisors}: 
\begin{itemize}
	\item[*] Venus V. (Department of Natural Resource, ITC)
	\item[*] Toxopeus A.G. (Department of Natural Resource, ITC)
\end{itemize}

\begin{figure}[h]
	\begin{center}
	\includegraphics[height=2cm,width=8cm]{ITC-logo.jpg}
	\end{center}
\end{figure}
\pagebreak

\section*{}
	\markright{Table of Content}
	\addcontentsline{toc}{section}{Table of Content}
	\tableofcontents 
\pagebreak

\section*{Acknowledgements}
	\markright{Acknowledgements}
	\addcontentsline{toc}{section}{Acknowledgements}
\pagebreak

\section*{Disclaimer}
	\markright{Disclaimer}
	\addcontentsline{toc}{section}{Disclaimer}
		This document describes work undertaken as part of a programme of study at the University of
		Twente, Faculty ITC. All views and opinions expressed therein remain the sole responsibility of
		the author, and do not necessarily represent those of the university.\\
\pagebreak

\section*{Abstract}
	\markright{Abstract}
	\addcontentsline{toc}{section}{Abstract}
\pagebreak

\section*{List of Tools}\label{sec:0}
	\markright{List of Tools}
	\addcontentsline{toc}{section}{List of Tools}

	\textbf{Software and utilities used in the current work}:\\
	\begin{itemize}
		\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
		\item \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and 					classification
		\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water 					properties under changing environment
		\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and GPS for data capture during the fieldwork
		\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
		\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
		\item \href{http://www.python.org/}{Python language}, for writing auxiliary scripts (e.g. raw data interpolation)
		\item \href{http://www.gnuplot.info/}{Gnuplot} for plotting graphs
		\item \href{http://www.openoffice.org/}{Open Office} : document typesetting, data preprocessing and preliminary statistical evalution
		\item \href{http://calc2latex.sourceforge.net/}{Calc2LaTeX} : converting spreadsheets into \LaTeX environment
		\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)} : statistical data edition and analysis 
		\item \href{http://gretl.sourceforge.net/}{Gretl: Gnu Regression, Econometrics and Time-series Library} for statistic analysis
		\item \href{http://staruml.sourceforge.net/en/}{StarUML} , \href{http://live.gnome.org/Dia}{Dia} , \href{http://argouml.tigris.org/}						{ArgoUML}: diagramming flowcharts for work algorithms
		\item \href{http://inkscape.org/}{Inkscape}: drawing
		\item \href{http://www.notepad-plus-plus.org/}{Notepad++}: auxiliary text editing and raw data formatting
	\end{itemize}
\pagebreak

@@

\subsection{Research approach}
Seagrass consistent monitoring and mappint is necessary and important for the sustainable coastal
development and conservation measures. Earlier, many seagrass meadows have been destroyed by
human activities in the coastal zone, mainly due to the ignorance of their existence, because
information on the seagrass bed exact location was not available \cite{Choo06}\label{Choo06}. Well-time seagrass
observations and mapping enables precise control of its spatial distribution, detection of any changes
in the seagrass landscapes, highlightes potential environmental threats in the coastal zone (e.g.
declining of meadows) before they become unmanageable for the coastal management services.
Choosing the right and most effective approach method for the seagrass monitoring is essential.
Remote sensing methods alone, though having evident advantages, are insufficient, because satellite
images of underwater habitats are notoriously difficult to identify and interpret. The best research
method should be based on the integrated approach, well described in various scientific works \cite{Brown02,Montefalcone06,Kirkman96}\label{Brown02} \label{Montefalcone06} \label{Kirkman96}, which includes combination of various
techniques of the seagrass monitoring, i.e. remote sensing imagery classification of aerial and satellite
images, GIS-based spatial analysis and ground in-situ surveys.
The current study is based on the application of the remote sensing data, broadband satellite imagery,
aerial images and the results of the underwater videographic measurements towards seagrass
mapping  (Fig.\ref{fig:1.11}). Image classification is based on the principle of the differentiation between the spectral
signatures of various seafloor cover types  (Fig.\ref{fig:1.10}). The spectrum of light coming up from the ocean surface in
shallow waters keeps information on the optical properties of the seawater components and benthic
substrate which can be read from their spectral signatures \cite{Werdell03}\label{Werdell03}. \\

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.48]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}}
	\label{fig:1.10}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.6]{UML_GIS.png}
	\caption{General methodological research approach. \href{http://inkscape.org/}{Inkscape}}
	\label{fig:1.11}
\end{figure}

@@

\subsection{Seagrass global monitoring: history and perspectives}
Mapping and monitoring the seagrass is important for the environmental assessment of the marine
ecosystemsn in coastal areas. Regular tracking of current distribution of seagrass meadows, based on
correct information and cartographic visualization of seagrasses, is a preventive environmental
management, which helps to analyse potential environmental risks of coastal areas, decrease of the
number of species, loss of meadows and patches of the seagrasses.
The tradition of global seagrass mapping though has not a very long history comparing to the
terrestrial cartography, due to the technical difficulties of underwater observations, yet nowadays is
has become a successful, rapidly developping, increasingly popular and challenging research branch.
Seagrasses can be mapped using various approaches, with the most popular method of the remote
sensing application, enabling to detect and discriminate the substratum and various vegetation types
growing on the seafloor.
Several global seagrass surveys organise and provide regular monitoring of the segrass distribution,
health and species sustainability, amongst which the most world-known are, perhaps, the following
ones: \href{http://www.seagrassnet.org/}{Global Seagrass Monitoring project}, \href{http://www.seagrasswatch.org/}{Seagrasswatch}, 
located in Australia, two European-located institutions - \href{http://mediterranean.seagrassonline.org/}{the Mediterranean association Seagrass-2000} and \href{http://www.imedea.uib.es/index.php}{the Mediterranean Institute for Advanced Studies}, US American 
\href{http://www2.fiu.edu/~seagrass/}{Seagrass Ecosystems Research Laboratory in South Florida} and
\href{http://www.flseagrass.org/}{Florida Seagrass organisation}, the \href{http://wsa.seagrassonline.org/}{World Seagrass Association} and  
the \href{http://www.seagrasses.org/}{Seagrasses.org}.
 These organizations aim at the global seagrass monitoring, providing with
research results and reporting guidelines and manuals with standartized methods and
recommendations, specific for the seagrass research and monitoring. There are also university marine
centres and research institutes conducting seagrass monitoring and as a particular part of their
research. Their reports and guidelines were used for references in the current research.
Regular obserbvations and monitoring of the seagrasses are known since 1960s, mainly in tropical
regions (Australia). Since that time traditional methods of the seagrass monitoring and common
recommendations are being elaborated. The development of the underwater SCUBA diving
equipment and devices enabled to conduct underwater detailed measurements and observations
largely contributed to the improvement of the traditional in-situ observations of seagrass. From the
other side, development of the remote sensing methods and data acquisition from space contributed to
the new methods of seagrass mapping, using distance approach and generally based on images
classification.

@@

\begin{figure}
	\centering
	\subfloat [RAMSES-ACC-UV - Hyperspectral UVA/UVB Irradiance Sensor: 280-500 nm]{\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-2-1.jpg}}
	\subfloat [RAMSES-ARC - Hyperspectral UV-VIS Radiance Sensor: 320-950 nm.]{\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-2-2.jpg}}
	\caption{RAMSES Hyperspectral Sensors. Source: \href{http://www.trios.de}{Trios}}
	\label{fig:2.1}
\end{figure}

The optical properties of the sea water vary with different environmental conditions and reflect
current chemical content and physical specifics of the water, revealing the variability and distribution
of color of the sea waters, determined by the material in the water, e.g. chlorophyll, as well as its
physical properties, e.g. water absorption, attenuation, backscattering \cite{Maffione01a}\label{Maffione01a}.
Shallow waters generally contain more dissolved substances and suspended particles, which directly
influnces the transparency and colour of the waters of shelf zones \cite{Jerlov52}\label{Jerlov52}. Being highly
dynamic environments, coastal waters experience a variety of processes which alter their optical
properties incessantly. The effects of these processes influence application of the hyperspectral
remote sensing and reinforce other proceses \cite{Maffione01b}\label{Maffione01b}. Thus, waves and tides, increase
sedimentation processes which in turn may change microrelief properties and topology. 
The optical properties of the water are best reflected in the values of its radiance and irradiance,
which can be converted into spectral reflectance, or reflectivity. The spectral irradiance (E) is a
radiant flux of the electromagnetic solar radiation energy, received per surface unit area in a given
time (W·m−2·nm−1), while radiance (L) characterizes total emission or reflection that passes through
or is emitted from a particular area (W·sr−1·m−2 ·nm−1). Therefore, the spectral reflectance, or the
reflectivity of the object, can be estimated by the direct mathematical division of these first two
characteristics, and is expressed in persentage.
Therefore, the irradiance and radiance of the water should be measured in order to estimate spectral
reflectance of the various seafloor cover types. The optical measurements of the irradiance and
radiance of the sea water and bottom cover types of the seafloor can be received by the means of the
optical sensors spectroradiometers. There are several companies producing radiometers with various
characteristics and adjusted for defferent purposes, e.g. portable and miniature spectrometers from
\href{http://www.stellarnet.us}{StellarNet}, Hyperspectral Ocean Colour Radiometer (HyperOCR) sensor by
\href{http://www.satlantic.com}{Satlantic}, traceable spectroradiometers by \href{http://www.orboptronix.com}{Orbotronix} 
and lots of others. Among other radiometers there are ones designed by the \href{http://www.trios.de/}{Trios company} producing
optical sensors, GER-Series Field portable spectroradiometers from
\href{http://www.spectrapartners.nl/}{SpectraPartners}, etc. Trios-RAMSES hyperspectral radiometers
(Trios-RAMSES Hyperspectral UVA/UVB Irradiance Sensor and RAMSES-ARC Hyperspectral UV-
VIS Radiance Sensor) are small-sized, low power-consumpting, flexible for fieldwork yet with high
level of precision, specially calibrated for air and for water application as well as colour
measurements  (Fig.\ref{fig:2.1}). These products have been used for the radiance and irradiance
measurements in Agia Pelagia bay, Crete island, 2009.

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-12.jpg}
	\caption{Difference between broaddand multispectral and hyperspectral resolution of
		spectral signatures. Source: \cite{Clark99}\label{Clark99}}
	\label{fig:2.3}
\end{wrapfigure}

Seagrass meadows may reach spatial scales from several up to hundreds of metres, therefore they are
susceptible by the means of satellite imagery from remote sensors, both with moderate resolution
(e.g., \href{http://landsat.gsfc.nasa.gov/}{Landsat MSS, Landsat TM, Landsat ETM+}, \href{http://envisat.esa.int/instruments/meris/}{MERIS}, \href{http://asterweb.jpl.nasa.gov/}{ASTER}, \href{http://modis.gsfc.nasa.gov/}{MODIS}) and high resolution as
well (e.g., \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird}, \href{http://www.spotimage.fr/}{SPOT}, \href{http://www.itres.com/products/imagers/casi550}{CASI}). The possibility of their application towards seagrass mapping
vary and is limited by the technical characteristics (Table \ref{tab:1}) and resolution of these sensors  (Fig.\ref{fig:2.3}). In the next subparagraphs
we briefly discuss the limitations and research experience of the using of various imagery
for the seagrass mapping.

@@

\subsubsection{Review of multispectral imagery used for seagrass mapping}
Seagrass mapping using remotely sensed data from multispectral sensors is based on the classification
and discrimination of the seafloor types using their spectral characteristics in different wavebands.
Perhaps, the most known imagery, widely used in the remote sensing mapping, is received from the
group of Landsat sensors, known for its historical, pioneeer role in the satellite industry.
Being the longest running satellite system, lunched in 1972, Landsat is the only source of archival
data going back to 1984 at a sufficient spatial resolution \cite{Dekker05b}\label{Dekker05b}, which makes its data
desirable for historial mapping or environmental analysis of change detection of the seagrass
landscapes.
The Landsat TM and Landsat ETM+ data, with the most recent from sensor Landsat 7, an advanced
and multispectral scanning, launched in 1999, proved to be feasibile and useful for the mapping of
submerged vegetation, such as seagrasses or coral reefs. The successful applications of the imagery
Landsat TM towards the seagrass mapping, were reported in numerous research works \cite{Palandro03,Gullstroom06,Wabnitz08,Bierwirth93,Ferguson97,Rasib97}\label{Palandro03}\label{Gullstroom06}\label{Wabnitz08}\label{Bierwirth93}\label{Ferguson97}\label{Rasib97}.\\ The Landsat data are particularly suitable for the
case of change detection of seagrass landscapes at a decadal scale, because being the main sensor
onboard the Landsat satellites, the Thematic Mapper (TM) provides the longest time series
available for change detection analysis over submerged vegetation \cite{Palandro03}\label{Palandro03}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-13.jpg}
 	\caption{Band coverage of ASTER and Landsat channels on the e/m spectrum. Source:\cite{Kalinowski04}\label{Kalinowski04}}
	\label{fig:2.4}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-14.jpg}
	\caption{Multispectral vs. hyperspectral band coverage. Source: \cite{Sabins97}\label{Sabins97}}
	\label{fig:2.5}
\end{wrapfigure}

Advanced Spectrometer for Thermal Emission and Reflection Radiometer (\href{http://asterweb.jpl.nasa.gov/}{ASTER}), launched in
1999 onboard Terra sensor, provides high-resolution images of the Earth in 15 different bands of the
electromagnetic spectrum, ranging from visible to thermal infrared light  (Fig.\ref{fig:2.4}). It has diverse
subsystems for the visible near infrared (VNIR) with 15 m resolution, shortwave infrared (SWIR), and
thermal infrared (TIR) wavelength regions \cite{Kalinowski04}\label{Kalinowski04}. For each channel there is separate
onboard calibration (OBC) system, telescope with independent pointing and different detector technsology \cite{Arai05}\label{Arai05}. 
This makes ASTER imagery expecially suitable for the application towards detailed
mapping of surface temperature, emissivity and reflectance of objects and bathymetric elevations as
well. Successful application of the ASTER imagery towards mapping of submerged vegetation
reported, for example, by Hirose \cite{Hirose04} \label{Hirose04}. Other multispectra images have also been used for the
seagrass interpretation. For example, the success of using imagery from the multispectral airborne
scanner Daedalus AADS1268 is reported by Heege et al \cite{Heege03}\label{Heege03} where they aim at classification of
macrophytes in shallow waters of the Lake Constance.
In regard to the methods chosen for the image interpretation, the supervised classifications proves to
be the most worthy in a majority of case studies \cite{Palandro03,Peneva08}\label{Palandro03} \label{Peneva08}. While
often methods of the unsupervised classification are used as a tool for classifying submerged object
and features on the multispectral images and aerial imagery \cite{Fletcher09}\label{Fletcher09}, it still does not
provide common algorithms that can be applied to other images and regions. Thus, the strength of the
optical signal coming from the various types of the seafloor is strongly influenced by the effects of the
water column, its depth and chemical properties. Yet methods of the unsupervised classification
overleap mixed spectral effects of the water column which shift the real values of the spectra, with the
pure reflectance from the benthos; as a result, it may cause significant errors \cite{Dierssen03}\label{Dierssen03}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-15.jpg}
	\caption{Reflectance spectra of seagrass; Rrs(0-) - subsurface RS reflectance; Rrs b - the bottom
		reflectance. Source:\cite{Yang10}\label{Yang10}}
	\label{fig:2.6}
\end{wrapfigure}

Therefore, for the accurate assessments of various seafloor cover types, water
column depth and its optical properties, methods of supervised classification should be preferably
used as being more suitable in classification and interpretations of imagery.
While this and other studies \cite{Phinn08}\label{Phinn08}demonstrated the advantages and success of the
application of multispectral imagery for the spectral discrimination of the seafloor cover types and
mapping the submerged landscapes on the basis of pixels classification, the application of data from
hyperspectral sensors has better potential due to their higher resolution (Fig.\ref{fig:2.5}).
